---
title: "Session 1: Load and explore movie data"
author: "Emily Myers"
date: 03/23/2026
format: html
editor: visual
---

Throughout this workshop series, we are trying to answer two questions using a dataset on movies:

-   **A:** (How) do critical and popular opinions about movies diverge?

-   **B:** What genre of movies has the most bang for its buck (lowest budget, highest revenue?)

Before we can answer those questions, we need to explore our data!

# Load packages and data

## (Install) and load packages

Packages are groups of commands/functions that you can run in R. Each package has documentation in the “help” section of R (bottom right quadrant) and on CRAN.

We will use two packages:

`tidyverse` (actually a group of packages!) commonly used for data cleaning, plotting, etc.

`here` tells R which project folder to work in

If you have used these packages before, you do not need to re-install them (hint: look in the "packages" tab of the bottom right quadrant of R studio- if they show up when you search, they are installed!). If this is your first time using these packages, you need to install them in this code chunk:

```{r}
install.packages("tidyverse")
install.packages("here")
```

Regardless of whether or not you needed to install these packages, you will need to load them to use them in this session:

```{r}
library(tidyverse)
library(here)
```

### Load data

Now, we need to load our data into our global environment.

```{r}
movies <- read.csv(here("data", "movie_data.csv"))

```

# Explore data

Now that we've loaded our data, we can start exploring it! We'll start by looking at the structure and content of our data, and then dive into a more granular exploration of the variables most relevant to our questions. Our goals are two-fold:

1.  Identify necessary cleaning and wrangling steps
2.  Figure out if our data is suited to answer our questions

## Examine structure and content of the data

### Print the first few rows

```{r}
head(movies)
```

### Get the column names

```{r}
colnames(movies)
```

### Examine variable types

```{r}
lapply(movies, class)
```

### Accomplish all of this with one command!

```{r}
str(movies)
```

### Look for NAs in each column

```{r}
movies |>
  summarize(across(everything(), ~ sum(is.na(.))))
```

There are a lot of NAs in `meta_score`!

## Dig into most relevant variables

### Get summary statistics for all numeric variables

```{r}
summary(movies)
```

### Get summary statistics for variables relevant to question A (Rating, meta_score)

```{r}
summary(movies[, c("Rating", "meta_score")])
```

### Get summary statistics for variables relevant to question B (budget_value, grossww_value)

```{r}
summary(movies[, c("budget_value", "grossww_value")])

```

### Examine genres variable

How should we explore genres (also relevant to question B), since it isn't a numeric variable?

```{r}
# how many unique values?
length(unique(movies$genres))
# frequency of the top 20 genres
movies %>% count(genres, sort = TRUE) %>% slice_head(n = 20)
```

## Create simple visualizations of key variables

Here, we are using base R and making very simple plots. Later, you’ll learn how to make customizable (aka pretty) visualizations using `ggplot`.

### Histogram of Rating

```{r}
hist(movies$Rating)
```

### Histogram of meta_score

```{r}
hist(movies$meta_score)
```

### Histogram of budget_value

```{r}
hist(movies$budget_value)
```

Scientific notation makes things hard to interpret! Let's fix that and try again:

```{r}
options(scipen = 999)
hist(movies$budget_value)
```

### Histogram of grossww_value

```{r}
hist(movies$grossww_value)
```

# What did we learn about our data?

## What cleaning steps might we need to take?

-   standardize column names

-   clean/wrangle the genres variable

-   clean the budget column- should there be a movie with a \$1 budget?

-   make the budget and gross value variables more manageable

## Can we answer our questions with the data as is?

-   we have some NAs in meta score: for question A, we'll need to filter out rows with NA in meta score

-   we need a variable that measures "bang for buck"

-   the genres variable is not workable as is!
