---
title: "Wrangle and Analyze Film Data"
author: "Jeremy Mikecz"
format: html
editor: visual
---

## Questions

1.  From a practical research (not computational) perspective: How would/could you go about answering the question: what film genres offer the best return on investment?

```         

```

2, With your team, as you continue exploring the film dataset, what other question would you like to answer?

```         


```

```{r}

```

```{r}

```

```{r}

```

# Introduction to Tidyverse

Tidyverse is a system of R packages for data wrangling and analysis. It provides a different method and syntax for working with data tables (often called "data frames" in data science; but, tidyverse dataframes are known as "tibbles") from base R.

For additional help learning how to use R's tidyverse system of packages, see:

-   Wickham, Çetinkaya-Rundel, and Grolemund, [*R for Data Science*](https://r4ds.hadley.nz/)

-   Grolemund, [A Tidyverse Cookbook](https://rstudio-education.github.io/tidyverse-cookbook/) (2020).!

-   package documentation: [tidyverse](https://tidyverse.tidyverse.org/)

-   [tidyr cheatsheet](https://github.com/rstudio/cheatsheets/blob/main/tidyr.pdf)

-   [*Tidyverse style guide*](https://style.tidyverse.org/)

For a more detailed introduction to working with tidy data in R see:

-   Wickham, Çetinkaya-Rundel, and Grolemund, *R for Data Science*, Ch. 5 ["Data Tidying"](https://r4ds.hadley.nz/data-tidy).

-   Grolemund, [A Tidyverse Cookbook](https://rstudio-education.github.io/tidyverse-cookbook/) (2020).

-   Silge and Robinson, [Text Mining with R](https://www.tidytextmining.com/), Ch. 1 "The tidy text format".

## Getting Started with Tidyverse

Data analysis usually involves working with two-dimensional datasets known as dataframes. There are different ways to work with dataframes in R, these include:

-   using core R dataframe functions

-   using the **tibble** a type of dataframe used with the tidyverse collection of packages

-   using other dataframe types like **data.table** for speed and to facilitate working with large datasets (i.e. with millions of rows)

In these lessons, we will work with [tidyverse](https://www.tidyverse.org/) a collection of packages designed for data science. Tidyverse works with **tibbles**, a customized and newer form of dataframes. For more on the differences between tibbles and dataframes see the [explanation here](https://r4ds.had.co.nz/tibbles.html). Key packages in tidyverse include (borrowing from Kyle Walker's [**Analyzing US Census Data: Methods, Maps, and Models in R**](https://walker-data.com/census-r/index.html)**)**:

```         
* readr (Wickham and Hester 2021), contains tools for importing and exporting datasets;

* dplyr (Wickham et al. 2021), a powerful framework for data wrangling tasks;

* tidyr (Wickham 2021b), a package for reshaping data;

* purrr (Henry and Wickham 2020), a comprehensive framework for functional programming and iteration;

* ggplot2 (Wickham 2016), a data visualization package based on the Grammar of Graphics
```

```{r}
## install tidyverse with:
#install.packages("tidyverse")

## update tidyverse with:
#tidyverse_update()

## import tidyverse with:
library(tidyverse)

## see what packages are included with tidyverse:
tidyverse_packages()
```

## Combine functions into a pipe

Using the symbol `|>` we can chain multiple functions together.

::: callout-note
Previously, `%>%` was commonly used for pipes. Now, it is recommended to use `|>` instead. In R Studio, you can use the keyboard shortcut `CTRL/CMD + SHIFT + M` to create a pipe. However, you can update R Studio's default pipe using the instructions in *R for Data Science*, [Ch. 3.4](https://r4ds.hadley.nz/data-transform.html#sec-the-pipe).
:::

Each new function in the pipeline operates on the results produced by the previous function.

Chaining functions together in a pipe like this:

```         
starwars |>
  group_by(species) |>
  summarise(avg_height = mean(height, na.rm = TRUE)) |>
  arrange(avg_height)
```

produces the same results as nesting a series of functions within another:

```         
arrange(
  summarise(
    group_by(starwars, species), 
    avg_height = mean(height, na.rm = TRUE)
  ), 
  avg_height
)
```

or calling each function in order:

```         
x1 <- starwars 
x2 <- group_by(x1, species)
x3 <- summarise(x3, avg_height = mean(height, na.rm = TRUE))
arrange(x3, avg_height)
```

By default `|>` passes the result of the left hand side to the the first unnamed argument of the function on the right hand side. To override this default, use `_` as a placeholder within the function call on the right hand side. `|>` will evaluate `_` as the result of the left hand side, instead of passing the result to the first unnamed argument.

```         
starwars |>
  lm(mass ~ height, data = _)

# the old piping syntax
#starwars %>%
#  lm(mass ~ height, data = .)
```

## Tidy Data

From the [Tidyverse cookbook](https://rstudio-education.github.io/tidyverse-cookbook/tidy.html):

Data tidying refers to reshaping your data into a tidy data frame or [tibble](https://rstudio-education.github.io/tidyverse-cookbook/how-to-use-this-book.html#tidy-data). Data tidying is an important first step for your analysis because every tidyverse function will expect your data to be stored as **Tidy Data**.

Tidy data is tabular data organized so that:

1.  Each column contains a single variable

2.  Each row contains a single observation

Tidy data is not an arbitrary requirement of the tidyverse; it is the ideal data format for doing data science with R. Tidy data makes it easy to extract every value of a variable to build a plot or to compute a summary statistic. Tidy data also makes it easy to compute new variables; when your data is tidy, you can rely on R's rowwise operations to maintain the integrity of your observations. Moreover, R can directly manipulate tidy data with R's fast, built-in vectorised observations, which lets your code run as fast as possible.

The definition of Tidy Data isn't complete until you define variable and observation, so let's borrow two definitions from [*R for Data Science*](https://r4ds.had.co.nz/exploratory-data-analysis.html):

1.  A **variable** is a quantity, quality, or property that you can measure.

2.  An **observation** is a set of measurements made under similar conditions (you usually make all of the measurements in an observation at the same time and on the same object).

As you work with data, you will be surprised to realize that what is a variable (or observation) will depend less on the data itself and more on what you are trying to do with it. With enough mental flexibility, you can consider anything to be a variable. However, some variables will be more useful than others for any specific task. In general, if you can formulate your task as an equation (math or code that contains an equals sign), the most useful variables will be the names in the equation.

### Create a tibble from scratch

Rarely will you ever create a dataframe or tibble from scratch except to create small practice datasets.

Here, we create a small practice dataset about the popular murder mystery TV show, [*Only Murders in the Building*](https://www.imdb.com/title/tt11691774/)

```{r}
omitb <- tribble(~name, ~occupation, ~apartment, ~is_suspect, ~is_dead,             "Charles", "washed-up actor", "14C", "no", "no",             "Mabel", "unemployed never-was", "12E", "maybe", "no",             "Oliver", "theater director", "10D", "no", "no",             "Howard", "childless cat guy", "3D", "yes", "no",             "Bunny", "petty despot", "12A", "no", "maybe" )
omitb
```

We can add a new row by piping the function **add_row.** To figure out how do so, we can review the [function's documentation](https://tibble.tidyverse.org/reference/add_row.html). Note: the default is to add the new row to the bottom of the tibble. To add it before the first row we need to set the argument `.before = 1`.

```{r}
omitb <- omitb |>
  add_row(name="Jan", occupation = "2nd-chair basoonist", 
          is_suspect="yes", is_dead="no", .before=1)

omitb
```

Above we used the **tribble** function to create a **tibble** (tribble vs. tibble - confusing right?) by using a syntax that lines up the rows neatly in a way that will preview the result.

You can also create the same tibble using the **tibble** function and passing in a list with the values of column:

```{r}
tibble(name = c("Charles", "Mabel", "Oliver", "Howard", "Bunny"),
       occupation = c("washed-up actor", "unemployed never-was", "theater director", "childless cat guy", "petty despot"),
       apartment = c("14C", "12E", "10D", "3D", "12A"),
       is_suspect = c("no", "no", "no", "yes", "no"),
       is_dead = c("no", "no", "no", "no", "maybe"),
       )
```

## Dataframes vs. tibbles

Read more about tibbles in *R for Data Science*, [Ch. 10](https://r4ds.had.co.nz/tibbles.html), in the [tidyverse documentation](https://tibble.tidyverse.org/),

```{r}
library(here)
df <- read.csv(here("data/movie_data.csv"))
print(class(df))

# convert a dataframe to a tibble with as_tibble:
df2 <- as_tibble(df)
print(class(df2))
```

```{r}
df <- read_csv(here("data/movie_data.csv"))
class(df)
```

## 3. Tidyverse / dplyr verbs for Data Wrangling

To modify and "wrangle" datasets, the tidyverse commonly uses the following verbs / functions:

-   **select()**: to select specific columns by their names or data types

-   **arrange()**: to order rows by one or more columns

-   **rename()**: to rename columns

-   **mutate()**: to create columns

-   **filter()**: to filter out rows by a given condition

-   **distinct()**: to keep only distinct / unique rows

-   **gather()**: to make "wide" data longer

-   **spread()**: to make "long" data wider

-   **separate()**: to split a single column into multiple columns

-   **unite()**: to combine multiple columns into one

[More tidyverse verbs and functions here](https://dplyr.tidyverse.org/reference/index.html).

# Common Data Science Operations in Tidyverse

+------------------------------------+------------------------------------------+----------------------------+---------------------------------+
| Task                               | Python - Pandas                          | Base R                     | Tidyverse (R)                   |
+====================================+==========================================+============================+=================================+
| **Sort table by column**           | df.sort_values('col')                    | sort(df\$col)              | df \|\>                         |
|                                    |                                          |                            |                                 |
|                                    |                                          |                            | arrange(col)                    |
+------------------------------------+------------------------------------------+----------------------------+---------------------------------+
| **Filter rows**                    | df.loc\[df\['col'\]\>0,:\]               | df\[df\[col\] \> 0, \]     | df \|\>                         |
|                                    |                                          |                            |                                 |
|                                    |                                          |                            | filter(col \> 0)                |
+------------------------------------+------------------------------------------+----------------------------+---------------------------------+
| **Subset columns**                 |                                          |                            | df \|\>                         |
|                                    |                                          |                            |                                 |
|                                    |                                          |                            | select(col)                     |
+------------------------------------+------------------------------------------+----------------------------+---------------------------------+
| **Create a new column**            | df.loc\[:, "square"\] = df\[col\] \*\* 2 | df\$square = df\[col\] \^2 | df %\>%                         |
|                                    |                                          |                            |                                 |
|                                    |                                          |                            | mutate(square = col \^ 2)       |
+------------------------------------+------------------------------------------+----------------------------+---------------------------------+
| **Perform calculation on column**  | sum(df\["col"\])                         | sum(df\$col)               | df \|\>                         |
|                                    |                                          |                            |                                 |
|                                    |                                          |                            | summarise(sumcol = sum(col))    |
+------------------------------------+------------------------------------------+----------------------------+---------------------------------+
| **Delete duplicate rows**          | df.drop_duplicates()                     | df\[!duplicated(df), \]    | df \|\>                         |
|                                    |                                          |                            |                                 |
|                                    |                                          |                            | distinct()                      |
+------------------------------------+------------------------------------------+----------------------------+---------------------------------+
| **Group/Apply/Combine**            | df.groupby("col").sum()                  | by(df, col, sum(x))        | df \|\>                         |
|                                    |                                          |                            |                                 |
|                                    |                                          | ???                        | group_by(col) \|\>              |
|                                    |                                          |                            |                                 |
|                                    |                                          |                            | summarise(groupsum = sum(col2)) |
+------------------------------------+------------------------------------------+----------------------------+---------------------------------+
| **Reshape data from wide to long** | *pd.wide_to_long()*                      | *melt()*                   | *use pivot_longer()*            |
+------------------------------------+------------------------------------------+----------------------------+---------------------------------+

## rename(): Rename columns

```{r}
colnames(df)
```

```{r}
df <- df |>
  rename(
    title=Title,
    mpa=MPA,
    rating = Rating,
    languages = Languages
  )
```

## select(): subset dataset by column name

```{r}

```

## Clean / Wrangle Scores

Visually review our dataframe `df` by opening it in R Studio (either double-click on `df` in the `Environment` pane to the top right or run `view(df)`).

```{r}
view(df)
```

## filter(): remove extreme or suspicious values

Try sorting different columns (by clicking on the column names). Do you notice any extreme values that are suspicious data?

Notice, for example, one movie has a budget of \$1. While there may be an explanation for this, let's remove it to avoid skewing some of our analysis:

```{r}
print(dim(df))
df <- df |>
  filter(budget_value >= 1000)
print(dim(df))
```

# Calculate Profit

## Make Numbers more manageable

```{r}
summary(df$grossww_value)
```

## mutate(): Visualize Distributions

Convert to millions:

```{r}
df <- df |> 
  mutate(budget_mm = budget_value / 1000000,
         grossww_mm = grossww_value / 1000000)
```

Calculate Profit

```{r}
df = df |>
  mutate(net_profit_mm = grossww_mm - budget_mm)
```

Calculate Profit as Percent of Budget

```{r}
df = df |>
  mutate(profit_pct = net_profit_mm /  budget_mm * 100)
```

### Export

```{r}
write_csv(df, here("data/movie_data_financial.csv"))
```

# Review Genres

Let's take a look at the genre column:

```{r}
df$parent_genres[0:10]
```

Each film appears once per genre, so multi-genre films are counted multiple times. This means genres that frequently co-occur may have correlated estimates. It also means a film with two assigned genres is counted just twice while one with five is counted five times.

Some possible solutions:

-   **one row per film/genre (EASIEST)**: split genres into different rows, so that a five-genre film is split out into five rows

-   **one row per film genre (weighted)**: split genres into different rows but weight each genre by its proportion. So a "horror-comedy" means horror counts for 0.5 genres and comedy also for 0.5.

-   **one row per film, one column per genre**\*\* (wide/dummy format): create a separate binary column (TRUE / FALSE) for each genre; best for statistical modeling and controlling (i.e. "controlling for other genres, what is horror's effect?")

Let's do the simplest approach: one row per film / genre (no weighting).

*whiteboard*

*demo string functions a bit*

```{r}

```

## separate_rows():

```{r}
genre_df <- df |>
  separate_rows(parent_genres, sep = ",") |>   # split on comma
  mutate(parent_genres = str_trim(parent_genres))     # clean up whitespace around each genre name
```

## Examine Distributions of Genres

```{r}

```

# Split-Apply-Combine

## group_by() + summarize(): Median ROI by genre

```{r}
genre_aggreg <- genre_df |>
  group_by(parent_genres) |>
  summarize(
    median_roi = median(profit_pct, na.rm = TRUE),
    n = n()
  ) |>
  arrange(desc(median_roi))
  
```

```{r}
genre_aggreg |>
  filter (n>=20)
```

**Export list of genres with ROI and n**

```{r}
genre_aggreg_alpha <- genre_aggreg |>
  arrange(parent_genres)

# write_csv(genre_aggreg_alpha, here("data/genre_list.csv"))

```

**Add spread to the summary** Median alone doesn't capture risk. A genre with high median ROI but huge variance is a very different bet than one with moderate but consistent returns:

```{r}
genre_df |>
  group_by(parent_genres) |>
  summarize(
    median_roi = median(profit_pct, na.rm = TRUE),
    iqr_roi    = IQR(profit_pct, na.rm = TRUE),   # interquartile range = spread
    n          = n()
  ) |>
  filter(n >= 50) |>
  arrange(desc(median_roi))
```

The IQR is a natural companion to the median and tells you about consistency — which genres are reliable bets vs. high-risk/high-reward.

```{r}
head(genre_df)
```

```{r}
colnames(genre_df)
```

```{r}
genre_df <- genre_df |>
  select(-any_of("genres")) |>
  select(title:meta_score, parent_genres, date:last_col())
```

```{r}
write_csv(genre_df, here("data/movies_split-by_parent-genre.csv"))
```

## Closing

## Experiment

1.  We asked your team to come up with your own research question you would like to answer using the film dataset. With the time remaining, please do the following:
    1.  Identify data wrangling steps you would need to complete to answer this question
    2.  create a new notebook and carry out these data wrangling steps.

## 

1.  
