---
title: "Wrangle and Analyze Film Data"
author: "Jeremy Mikecz"
format: html
editor: visual
---

## Questions

1.  What

## interactive portions

```{r}
library(tidyverse)
```

```{r}
library(here)
df <- read_csv(here("data/movie_data_dollars.csv"))
```

```{r}
head(df)
```

Opener and Closer - referring back to big research questions

change filename - remove years and remove clean from the title of the file

parent genres

## Standardize Variable Names

```{r}
colnames(df)
```

```{r}
df <- df |>
  rename(
    title=Title,
    mpa=MPA,
    rating = Rating,
    languages = Languages
  )
```

## Clean / Wrangle Scores

Visually review our dataframe `df` by opening it in R Studio (either double-click on `df` in the `Environment` pane to the top right or run `view(df)`).

```{r}
view(df)
```

## remove extreme or suspicious values

Try sorting different columns (by clicking on the column names). Do you notice any extreme values that are suspicious data?

Notice, for example, one movie has a budget of \$1. While there may be an explanation for this, let's remove it to avoid skewing some of our analysis:

```{r}
print(dim(df))
df <- df |>
  filter(budget_value >= 1000)
print(dim(df))
```

# Calculate Profit

## Make Numbers more manageable

```{r}
summary(df$grossww_value)
```

## Visualize Distributions

Convert to millions:

```{r}
df <- df |> 
  mutate(budget_mm = budget_value / 1000000,
         grossww_mm = grossww_value / 1000000)
```

Calculate Profit

```{r}
df = df |>
  mutate(net_profit_mm = grossww_mm - budget_mm)
```

Calculate Profit as Percent of Budget

```{r}
df = df |>
  mutate(profit_pct = net_profit_mm /  budget_mm * 100)
```

### Export

```{r}
write_csv(df, here("data/movie_data_dollars2.csv"))
```

# Review Genres

Let's take a look at the genre column:

```{r}
df$genres[0:10]
```

Each film appears once per genre, so multi-genre films are counted multiple times. This means genres that frequently co-occur may have correlated estimates. It also means a film with two assigned genres is counted just twice while one with five is counted five times.

Some possible solutions:

-   **one row per film/genre (EASIEST)**: split genres into different rows, so that a five-genre film is split out into five rows

-   **one row per film genre (weighted)**: split genres into different rows but weight each genre by its proportion. So a "horror-comedy" means horror counts for 0.5 genres and comedy also for 0.5.

-   **one row per film, one column per genre**\*\* (wide/dummy format): create a separate binary column (TRUE / FALSE) for each genre; best for statistical modeling and controlling (i.e. "controlling for other genres, what is horror's effect?")

Let's do the simplest approach: one row per film / genre (no weighting).

*whiteboard*

*demo string functions a bit*

```{r}
genre_df <- df |>
  mutate(
    genre = genres |>
      str_remove_all("\\[|\\]|'") |>   # remove brackets and single quotes
      str_trim()                         # remove any leading/trailing whitespace
  ) 
```

```{r}
genre_df <- genre_df |>
  separate_rows(genre, sep = ",") |>   # split on comma
  mutate(genre = str_trim(genre))     # clean up whitespace around each genre name
```

## Map Parent Genres to Genres

```{r}
parent_genres <- read_csv(here("data/genre_list_edited_jm.csv"))
head(parent_genres)
```

```{r}
genre_df <- genre_df |>
  left_join(
    parent_genres |>
      select(genre, parent_genre),
    by = "genre"
  )
```

### Verify

```{r}
genre_df |> 
  filter(str_detect(genre, '"')) |> 
  select(genre) |> 
  head()
```

## Examine Distributions of Genres

```{r}

```

## Median ROI by genre

```{r}
genre_aggreg <- genre_df |>
  group_by(genre) |>
  summarize(
    median_roi = median(profit_pct, na.rm = TRUE),
    n = n()
  ) |>
  arrange(desc(median_roi))
  
```

```{r}
genre_aggreg |>
  filter (n>=20)
```

**Export list of genres with ROI and n**

```{r}
genre_aggreg_alpha <- genre_aggreg |>
  arrange(genre)

# write_csv(genre_aggreg_alpha, here("data/genre_list.csv"))

```

**Add spread to the summary** Median alone doesn't capture risk. A genre with high median ROI but huge variance is a very different bet than one with moderate but consistent returns:

```{r}
genre_df |>
  group_by(genre) |>
  summarize(
    median_roi = median(profit_pct, na.rm = TRUE),
    iqr_roi    = IQR(profit_pct, na.rm = TRUE),   # interquartile range = spread
    n          = n()
  ) |>
  filter(n >= 50) |>
  arrange(desc(median_roi))
```

The IQR is a natural companion to the median and tells you about consistency — which genres are reliable bets vs. high-risk/high-reward.

```{r}
head(genre_df)
```

## 

```{r}
colnames(genre_df)
```

```{r}
genre_df <- genre_df |>
  select(-any_of("genres")) |>
  select(title:meta_score, genre, parent_genre, date:last_col())
```

```{r}
write_csv(genre_df, here("data/movie_split-by_genre.csv"))
```
