---
title: "Intro to GGplot"
author: "Jeremy Mikecz"
format: html
editor: visual
---

## 

## 0. Setup

```{r}
library(tidyverse)
library(here)
```

## 1. Import Data

Now let's import this dataset:

```{r}
movies <- read_csv(here("data/movie_data_dollars2.csv"))
```

```{r}
colnames(movies)
```

```{r}

```

## 2. First Plot: two variables

### 2a. Basic plot

Let's first create a scatterplot of this dataset for one year (1982) and two variables (`lifeExp` = life expectancy and `gdpPercap` = gdpPercap).

```{r}
```

```{r}
# let's store the year we want to examine in variable 
# so we can easily change it later
ggplot(movies, aes(x=budget_mm, y=grossww_mm)) +
  geom_point()
```

### 2b. Adjust color & transparency (as constants)

[ggplot colors](https://sape.inf.usi.ch/quick-reference/ggplot2/colour)

```{r}
ggplot(movies, aes(x=budget_mm, y=grossww_mm)) +
  geom_point(color="steelblue4", alpha = 0.3, size=3)
```

### 2c. Add a smoothed line

```{r}
p <- ggplot(movies, aes(x=budget_mm, y=grossww_mm)) +
  geom_point(color="steelblue4", alpha = 0.3, size=3) +
  geom_smooth(color="purple")                      # , method='lm', se=False

p
```

### 2 - Extra: examine distributions more closely

```{r}
#install.packages("ggExtra")
library(ggExtra)

ggMarginal(p, type='histogram')

```

## Dealing with Small Budget Movies: Rescale Data

```{r}
library(scales)
log_p <- ggplot(movies, aes(x=budget_mm, y=grossww_mm)) +
  geom_point(color="steelblue4", alpha = 0.3, size=3) +
  geom_smooth(color="purple") +
  scale_x_log10(labels = label_dollar(suffix = "M")) +
  scale_y_log10(labels = label_dollar(suffix = "M")) +
  annotation_logticks()
  
log_p
```

-   colors - parent genres

-   smoothed_lines

-   facets?

-   theme_minimal? basic preset themes

-   clusters?

-   aggreg by genre:

    -   animate by year

    -   hover

    -   connected scatter plot
