---
title: "Intro to GGplot"
author: "Jeremy Mikecz"
format: html
editor: visual
---

## 

## 0. Setup

```{r}
library(tidyverse)
library(here)
```

## 1. Import Data

Now let's import this dataset:

```{r}
movies <- read_csv(here("data/movie_data_dollars2.csv"))
```

```{r}
colnames(movies)
```

```{r}

```

## 2. First Plot: two variables

### 2a. Basic plot

Let's first create a scatterplot of this dataset for one year (1982) and two variables (`lifeExp` = life expectancy and `gdpPercap` = gdpPercap).

```{r}
```

```{r}
# let's store the year we want to examine in variable 
# so we can easily change it later
ggplot(movies, aes(x=budget_mm, y=grossww_mm)) +
  geom_point()
```

### 2b. Adjust color & transparency (as constants)

[ggplot colors](https://sape.inf.usi.ch/quick-reference/ggplot2/colour)

```{r}
ggplot(movies, aes(x=budget_mm, y=grossww_mm)) +
  geom_point(color="steelblue4", alpha = 0.3, size=3)
```

### 2c. Adjust color as a variable

```{r}
ggplot(movies, aes(x=budget_mm, y=grossww_mm, color=mpa)) +
  geom_point(alpha = 0.3, size=3)
```

### 2c. Add a smoothed line

```{r}
p <- ggplot(movies, aes(x=budget_mm, y=grossww_mm)) +
  geom_point(color="steelblue4", alpha = 0.3, size=3) +
  geom_smooth(color="purple")                      # , method='lm', se=False

p
```

### 2 - Extra: examine distributions more closely

```{r}
#install.packages("ggExtra")
library(ggExtra)

ggMarginal(p, type='histogram')

```

## Dealing with Small Budget Movies: Rescale Data

```{r}
library(scales)
log_p <- ggplot(movies, aes(x=budget_mm, y=grossww_mm)) +
  geom_point(color="steelblue4", alpha = 0.3, size=3) +
  geom_smooth(color="purple") +
  scale_x_log10(labels = label_dollar(suffix = "M")) +
  scale_y_log10(labels = label_dollar(suffix = "M")) +
  annotation_logticks()
  
log_p
```

## 3. Working with Genres

```{r}
genre_df <- read_csv(here("data/movies_split-by_parent-genre.csv"))
```

```{r}
head(genre_df)
```

```{r}
genre2020_df <- genre_df |>
  filter(year >= 2020)
```

```{r}
ggplot(genre2020_df, aes(x=budget_mm, y=grossww_mm, color=parent_genres)) +
  geom_point(alpha = 0.3, size=3) +
  geom_smooth(color="purple") +
  scale_x_log10(labels = label_dollar(suffix = "M")) +
  scale_y_log10(labels = label_dollar(suffix = "M")) +
  annotation_logticks() 
  
```

```{r}
ggplot(genre2020_df, aes(x = budget_mm, y = grossww_mm)) +
  geom_point(alpha = 0.3, size = 3) +
  geom_smooth(color="purple") +
  scale_x_log10(labels = label_dollar(suffix = "M")) +
  scale_y_log10(labels = label_dollar(suffix = "M")) +
  annotation_logticks() +
  facet_wrap(~ parent_genres, ncol=2)
```

```{r}
colnames(genre2020_df)
```

```{r}
ggplot(genre2020_df, aes(x = budget_mm, y = grossww_mm)) +
  geom_point(alpha = 0.3, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  facet_wrap(~ parent_genres, ncol=3) +
  labs(
    title    = "Budget vs. Worldwide Gross by Genre",
    subtitle = "Dashed line = break-even",
    x        = "Budget ($ millions)",
    y        = "Worldwide Gross ($ millions)"
  )
```

## ROI Distribution by Genre (Box Plot)

A bar chart of medians hides the spread. Use a box plot to see the **full distribution** of ROI across genres, including variability and outliers.

```{r}
ggplot(genre2020_df, aes(x = reorder(parent_genres, profit_pct, median), y = profit_pct)) +
  geom_boxplot(outlier.alpha = 0.2) +
  coord_flip() +
  labs(
    title = "Distribution of ROI by Genre",
    x     = "Genre",
    y     = "ROI (%)"
  )
```

## High-Return Low-Budget Films (Scatter Plot, Interactive)

To directly highlight films that made a lot relative to a small budget, we can size or color points by `profit_pct`, and filter to a budget range of interest. Using **plotly** makes it easy to hover and identify specific films.

```{r}
# install.packages("plotly")
library(plotly)

low_budget <- genre2020_df |>
  filter(budget_mm <= 50, profit_pct > 0)

p <- ggplot(low_budget, aes(
    x      = budget_mm,
    y      = grossww_mm,
    color  = parent_genres,
    size   = profit_pct,
    text   = paste0(title, "\nROI: ", round(profit_pct), "%")
  )) +
  geom_point(alpha = 0.5) +
  labs(
    title = "Low-Budget Films (≤ $50M) by Worldwide Gross",
    x     = "Budget ($ millions)",
    y     = "Worldwide Gross ($ millions)"
  )

ggplotly(p, tooltip = "text")
```

-   colors - parent genres

-   smoothed_lines

-   facets?

-   theme_minimal? basic preset themes

-   clusters?

-   aggreg by genre:

    -   animate by year

    -   hover

    -   connected scatter plot
