---
title: "Other Visualizations"
author: "Katie Owers Bonner"
format: html
editor: visual
---

# Session 4: Additional plot types, setting aesthetics, and saving figures

# Load data & packages if needed

```{r}
library(tidyverse)
library(here)
```

```{r}
movies <- read_csv("../data/movie_data_financial.csv")
```

# Exploring Critic vs Public Opinion

Let's come back to another question from our first day: (**How) do critical and popular opinions about movies diverge?**

The relevant variables are:

-   rating: IMDB's popular vote score (note the PG/R/etc. rating is called "mpa" in this dataset)

-   meta_score: Weighted average of professional critic scores

Reminder of what these variables look like from our exploratory analysis:

```{r}
summary(movies[, c("rating", "meta_score")])
```

```{r}
ggplot(data =movies, mapping = aes(x = meta_score, y=rating)) +
  geom_point() + 
  labs(
    title = "Critic vs public movie ratings",
    x = "Critic Score (meta_score)",
    y = "Public Score (rating)"
    )
```

## Does this association vary by MPA Rating?

Let's use an option called `facet_wrap` that splits up our plot by a factor (categorical variable) in the data. Let's use the other kind of rating - MPA.

```{r}
ggplot(data =movies, mapping = aes(x = meta_score, y=rating)) +
  geom_point()  + 
  labs(
  x = "meta_score (critics)",
  y = "rating (the public)"
  ) +
  facet_wrap(~mpa)
```

## What about by genre?

Let's load in the file where each movie is split out by genre and they are categorized into parent genres:

```{r}
movies_genre <- read_csv("../data/movies_split-by_parent-genre.csv")

```

```{r}
movies_genre |>
  summarise(unique_parent_genres = n_distinct(parent_genres))

```

```{r}
table(movies_genre$parent_genres)
```

And now do a figure where we facet by parent_genre:

```{r}
ggplot(data =movies_genre, mapping = aes(x = meta_score, y=rating)) +
  geom_point() + 
  labs(
  x = "meta_score (critics)",
  y = "rating (the public)"
  )  +
  facet_wrap(~parent_genres)
```

## Does the number of votes affect the relationship between public and critical opinion?

-   **`Votes`** = Total user votes on IMDb.

```{r}
options(scipen=999)
ggplot(data = movies, mapping = aes(x=votes)) +      
  geom_histogram() 
```

That's a lot of movies with not a lot of votes! Let's look more into it:

```{r}
summary(movies$votes)
```

How do movies with \<10,000 votes compare to the general distribution?

```{r}
movies <- movies |>
  mutate(low_votes = ifelse(votes<10000,"Yes","No"))
```

```{r}
ggplot(data =movies, mapping = aes(x = meta_score, y=rating, 
                               color = as.factor(low_votes))) +
  geom_point() + 
  labs(
    title = "Critic vs public movie ratings",
    x = "Critic Score",
    y = "Public Score"
    ) 
```

Let's make two visual tweaks here - label the legend better and change the color palette:

```{r}
ggplot(data =movies, mapping = aes(x = meta_score, y=rating, 
                               color = as.factor(low_votes))) +
  geom_point() + 
  labs(
    title = "Critic vs public movie ratings",
    x = "Critic Score",
    y = "Public Score",
    color = "<10k votes?"
    ) +
  scale_color_manual(values = c("Yes" = "red", "No" = "blue")) 


```

What other steps might you take to answer this question about the public vs critical opinion of movies?

# Boxplots and violin plots

What if we wanted to look instead at the values of rating (the public score) by parent_genre?

```{r}
ggplot(movies_genre, mapping = aes(x = parent_genres, y = rating)) +
  geom_boxplot() + 
  theme(
  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )
```

Take a minute to explore the theme axis text commands - what happens when you change various values?

You can also sort the data so that the histograms are arranged according to their median value:

```{r}
ggplot(movies_genre, mapping = aes(x = fct_reorder(parent_genres, -rating, .fun=median), y = rating)) +      
  geom_boxplot() +       
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)   
        )
```

This conveys a lot of information! What might be missing?

Let's add points to show how many movies were in each boxplot as well as the summary values.

```{r}
ggplot(movies_genre, mapping = aes(x = fct_reorder(parent_genres, -rating, .fun=median), y = rating)) +
  geom_boxplot(fill = "darkgrey")  + 
  theme(
  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  ) +
  geom_jitter(color="blue", size=0.4, alpha=0.3)
```

What about another way of looking at this data? Violin plots show the shape of the data distribution, combining elements of histograms and the underlying points.

```{r}
ggplot(movies_genre, mapping = aes(x = parent_genres, y = rating)) +
  geom_violin() + 
  theme(
  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )
```

# Line plots and changing aesthetics

What if you were curious about the number of movies in each parent genre by year?

First, let's calculate those yearly summaries:

```{r}
genre_summary_data <- movies_genre |>
  group_by(year, parent_genres) |>
  summarise(
    count = n()
  )
```

```{r}
ggplot(genre_summary_data, aes(x=year, y=count, 
                               group = parent_genres, color = parent_genres)) +
  geom_line()
```

There are some interesting patterns here! What do you see?

There is also a lot of information here, and it's a little hard to interpret information about the genres themselves with so many lines and the current color scheme.

Another visual tool is highlighting a single (or few) line(s) of interest. Say we were interested in trends in comedies. We could make all the other lines grey and leave comedies red to show them in context while also highlighting the category of interest:

```{r}
ggplot(genre_summary_data, aes(x=year, y=count, group = parent_genres, color = parent_genres)) +
  geom_line(color = "grey", alpha = 0.5) + 
  geom_line(data = subset(genre_summary_data, parent_genres == "comedy"), color = "red")
```

## Change the color Palette

What if you wanted to keep all lines, but use a different color palette for your figure?

```{r}
ggplot(genre_summary_data, aes(x=year, y=count, group = parent_genres, color = parent_genres)) +      
  geom_line() +       
  scale_color_hue(h = c(180, 300))
```

Or to explicitly control the color of each line? Maybe you've selected a color scheme to use for each throughout your paper, or there's a common convention in your field that you want to use.

```{r}
ggplot(genre_summary_data, aes(x=year, y=count, group = parent_genres, color = parent_genres)) +      
  geom_line() +       
  scale_color_manual(values=c("#69b3a2", "purple", "pink","red", "orange", "black","darkgrey", "lightblue", "yellow","violet", "lavender",  "white","darkgreen", "green","orchid","cyan" ))
```

There are some good resources for your color adventures at <https://r-graph-gallery.com/ggplot2-color.html>!

## ggplot also built-in themes that change the underlying appearance of figures

Let's look at some histograms with various themes:

```{r}
ggplot(data = movies, mapping = aes(x=duration_minutes)) +      
  geom_histogram(binwidth=10) +      
  xlim(50,max(movies$duration_minutes))  
```

```{r}
ggplot(data = movies, mapping = aes(x=duration_minutes)) +      
  geom_histogram(binwidth=10) +      
  xlim(50,max(movies$duration_minutes)) +
  theme_classic()
```

```{r}
ggplot(data = movies, mapping = aes(x=duration_minutes)) +      
  geom_histogram(binwidth=10) +      
  xlim(50,max(movies$duration_minutes)) +
  theme_bw()
```

See built-in theme options here: <https://ggplot2.tidyverse.org/reference/ggtheme.html>

# Saving plots

`ggsave` is the function for saving ggplots. It defaults to saving the last plot displayed, at a default size and resolution depending on the filetype

```{r}
ggsave("../results/plot_default.png")
```

```{r}
ggsave("../results/plot_20x20cm.pdf", width = 20, height = 20, units = "cm")
```

```{r}
ggsave("../results/plot_highres.jpeg", dpi=600)
```

You can also assign `ggsave()` a specific plot to save:

```{r}
p <- ggplot(data = movies, mapping = aes(x=duration_minutes)) +      
  geom_histogram(binwidth=10, fill ="red") +      
  xlim(50,max(movies$duration_minutes)) +
  theme_bw()


ggsave("../results/plot_red.png", p)
```

# Reflection

-   What other questions do you have about this data that can be answered by a visual?

    -   What about movie duration vs rating?

-   How do you want to apply what you learned to your own data?
