---
title: "Other Visualizations"
author: "Katie Owers Bonner"
format: html
editor: visual
---

# Session 4: Additional visualizations

# ! material to add !

**Question A:** (How) do critical and popular opinions about movies diverge?

for one of them, can do more of an evolution

-   start wiht a line graphs - ok, maybe too cluttered

-   faceted graph or ridgeline plot

-   Q2: What about movie duration vs rating?

palettes?

**Publishing Visualizations (30 mins?)**

1.  exporting visualizations, refining them for publication

    1.  save at higher resolution, in different formats

    2.  ggsave

2.  legends?

## Load data & packages if needed

```{r}
library(tidyverse)
library(here)
```

```{r}
df <- read_csv("../data/movie_data_2010-2025_clean_dollars.csv")
```

## The Question

Let's come back to Question A from yesterday: (**How) do critical and popular opinions about movies diverge?**

Our relevant variables are:

-   Rating: IMDB's popular vote score

-   meta_score: Weighted average of professional critic scores

Reminder of what these variables look like from our exploratory analysis:

```{r}
summary(df[, c("Rating", "meta_score")])
```

```{r}
ggplot(data =df, mapping = aes(x = meta_score, y=Rating)) +
  geom_point()
```

## Create a dataset of the top 4 genres

```{r}
genre_df <- df |>
  mutate(
    genre = genres |>
      str_remove_all("\\[|\\]|'") |>   # remove brackets and single quotes
      str_trim()                         # remove any leading/trailing whitespace
  ) |>
  separate_rows(genre, sep = ",") |>   # split on comma
  mutate(genre = str_trim(genre))   # clean up whitespace around each genre name
  
```

```{r}
sort(table(genre_df$genre), decreasing = T)
```

```{r}
top_genre_df <- genre_df |>
  filter(genre %in% c("Drama", "Comedy","Thriller", "Action"))
```

## Facet_wrap by genre

```{r}
ggplot(data =top_genre_df, mapping = aes(x = meta_score, y=Rating)) +
  geom_point() +
  facet_wrap(~genre)
```

show adding trendline? JEremy did this

```{r}
ggplot(data =top_genre_df, mapping = aes(x = meta_score, y=Rating)) +
  geom_point() +
  facet_wrap(~genre) +
  geom_smooth(method = "lm", se = FALSE, color = "blue")
```

# Boxplot/violin plot

rating by genre

```{r}
ggplot(top_genre_df, mapping = aes(x = genre, y = Rating)) +
  geom_boxplot()
```

```{r}
ggplot(top_genre_df, mapping = aes(x = genre, y = Rating)) +
  geom_boxplot(fill = "grey") +
  theme_bw() +
  geom_jitter(color="blue", size=0.4, alpha=0.7)
```

maybe redo this by duratin

```{r}
ggplot(top_genre_df, mapping = aes(x = genre, y = Rating)) +
  geom_violin()
```

```{r}
ggplot(top_genre_df, mapping = aes(x = genre, y = Rating)) +
  geom_violin() +
  geom_jitter(color="blue", size=0.5, alpha=0.7)
```

# Line plot

Number of movies in each of the 4 genres by year?

```{r}
summary_data <- top_genre_df |>
  group_by(year, genre) |>
  summarise(
    count = n()
  )
```

```{r}
ggplot(summary_data, aes(x=year, y=count, group = genre, color = genre)) +
  geom_line()
```

Additions

-   total number of films per year

-   total \# of genres per year

# Barplot (incl stacked)

Percent of these movie 4 genres over time (of the total)?

## Stacked barchart of genre by year

## Duration

```{r}
ggplot(data = df, mapping = aes(x=duration_minutes)) +   
  geom_histogram(binwidth=10) +   
  xlim(50,max(df$duration_minutes))   +   
  theme_classic()
```

```{r}
ggplot(data = df, mapping = aes(x=duration_minutes)) +   
  geom_density() +   
  xlim(50,max(df$duration_minutes))   +   
  theme_classic()
```

## 

# Aesthetics: Set ggplot Theme and Color Palette

# Extra Material ?

Facet by Year or MPA?

```{r}
ggplot(data =df, mapping = aes(x = meta_score, y=Rating)) +
  geom_point() +
  facet_wrap(~MPA)
```

```{r}
ggplot(data =df, mapping = aes(x = meta_score, y=Rating)) +
  geom_point() +
  facet_wrap(~year)
```

## Paired violin - would require different data structure

`{r} ggplot(top_genre_df, mapping = aes(x= genre, y = ))}`

# 
