---
title: "Other plot types, customization, and exporting figures"
author: "Katie Owers Bonner"
format: html
editor: visual
---

# Session 4: Additional plot types, setting aesthetics, and saving figures

# Load data & packages if needed

```{r}
library(tidyverse)
library(here)
```

```{r}
movies <- read_csv("../data/movie_data_financial.csv")
```

# Scatterplots, revisited: Exploring Critic vs Public Opinion

Let's come back to another question from our first day: (**How) do critical and popular opinions about movies diverge?**

The relevant variables are:

-   rating: IMDB's popular vote score (note the PG/R/etc. rating is called "mpa" in this dataset)

-   meta_score: Weighted average of professional critic scores

Reminder of what these variables look like from our exploratory analysis:

```{r}
summary(movies[, c("rating", "meta_score")])
```

A few notes

-   Different scales

-   Missing values (NA = 332) in meta_score

-   Difference in central tendency (mean and median) values \~ 64 (public) vs 55 (critic) rating

## Let's graph it!

Let's do a quick scatterplot to see if anything jumps out at us. We'll create a plot and assign it to the object named 'p' (for plot - a common convention in working with ggplot):

```{r}
p <- ggplot(data = movies, 
       mapping = aes(x = meta_score, y=rating)) +
  geom_point() 

p
```

Let's fix the labels in case you, like me, can never remember which is which. Because we've already saved p, all we need to do is add the label text to it:

```{r}
p + 
  labs(
    title = "Critic vs public movie ratings",
    x = "Critic Score (meta_score)",
    y = "Public Score (rating)"
    )
```

So far, this isn't a very illuminating plot.

## Does this association vary by MPA Rating?

Let's use an option called `facet_wrap` that splits up our plot by a factor (categorical variable) in the data. Let's use the other kind of rating - MPA.

```{r}
p +
  facet_wrap(~mpa)
```

Note: Because we didn't overwrite p with the labels we added above, they aren't showing here. We're just adding the facets to our initial basic scatterplot.

## Does the number of votes affect the relationship between public and critical opinion?

-   **`Votes`** = Total user votes on IMDb.

```{r}
options(scipen=999)
ggplot(data = movies, mapping = aes(x=votes)) +      
  geom_histogram() +
  scale_x_continuous(labels = scales::label_comma()) # note the ability to add a thousands separator!
```

That's a lot of movies with not a lot of votes! Let's look more into it:

```{r}
summary(movies$votes)
```

How do movies with \<10,000 votes compare to the general distribution?

```{r}
movies <- movies |>
  mutate(low_votes = ifelse(votes<10000,"Yes","No"))
```

Let's add that low_votes as a factor determining the color of each data point:

```{r}
ggplot(data = movies, 
       mapping = aes(x = meta_score, y=rating, 
                               color = as.factor(low_votes))) +
  geom_point() + 
  labs(
    title = "Critic vs public movie ratings",
    x = "Critic Score",
    y = "Public Score"
    ) 
```

## Legends

This is the first legend we've gotten! ggplot is pretty good about automatically deciding when a legend is needed. Let's make a few tweaks to it - label it better, move it inside the plot, and reverse the Yes/No entries.

```{r}
ggplot(data =movies, mapping = aes(x = meta_score, y=rating, 
                               color = as.factor(low_votes))) +
  geom_point() + 
  labs(
    title = "Critic vs public movie ratings",
    x = "Critic Score",
    y = "Public Score",
    color = "<10k votes?" # label for the legend
    ) +
  theme(legend.position = c(0.85, 0.25)) + # can use left, right, top, or bottom to place it outside the figure on the corresponding side.
  guides(color = guide_legend(reverse = TRUE)) # flips the legend order (so 'Yes' is on top now) 
```

More on legend customization here: <https://r-graph-gallery.com/239-custom-layout-legend-ggplot2.html>

To make this a bit more visually informative, let's make two quick aesthetic tweaks: change the colors and make the points semi-transparent:

```{r}
ggplot(data =movies, mapping = aes(x = meta_score, y=rating, 
                               color = as.factor(low_votes))) +
  geom_point(alpha = 0.5) + # transparency
  labs(
    title = "Critic vs public movie ratings",
    x = "Critic Score",
    y = "Public Score",
    color = "<10k votes?" 
    ) +
  scale_color_manual(values = c("Yes" = "red", "No" = "blue")) + # manually setting the color for each value of low_votes
  theme(legend.position = c(0.8, 0.2)) + 
  guides(color = guide_legend(reverse = TRUE)) 
```

What other steps might you take to answer this question about the public vs critical opinion of movies? Is there anything you want to explore?

# Boxplots and violin plots

What if we wanted to look instead at the values of rating (the public score) by parent_genre?

Let's load the genre data:

```{r}
movies_genre <- read_csv("../data/movies_split-by_parent-genre.csv") 
```

And get a quick reminder of what's in there:

```{r}
movies_genre |>    
  count(parent_genres, sort = TRUE)
```

Now let's plot the rating by genre:

```{r}
box <- ggplot(movies_genre, mapping = aes(x = parent_genres, y = rating)) +
  geom_boxplot() 

box
```

You can't read those parent_genre labels - let's fix that:

```{r}
box + 
  theme(
  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )
```

Take a minute to explore the `theme(axis.text.x = ...)` arguments - what happens when you change various values?

You can also sort the data so that the histograms are arranged according to their median value:

```{r}
box_sort <- ggplot(movies_genre, mapping = aes(x = fct_reorder(parent_genres, -rating, median), y = rating)) +      
  geom_boxplot() +       
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)   
        )

box_sort
```

This conveys a lot of information! What might be missing?

Let's add points to show how many movies were in each boxplot as well as the summary values.

```{r}
box_sort +
  geom_jitter(color="blue", size=0.4, alpha=0.3)
```

What about another way of looking at this data? Violin plots show the shape of the data distribution. They're actually vertically mirrored histograms!

```{r}
ggplot(movies_genre, mapping = aes(x = parent_genres, y = rating)) +
  geom_violin() + 
  theme(
  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )
```

# Line plots

What if you were curious about the number of movies in each parent genre by year?

First, let's calculate those yearly summaries:

```{r}
genre_summary_data <- movies_genre |>
  group_by(year, parent_genres) |>
  summarise(
    count = n()
  )
```

```{r}
ggplot(genre_summary_data, aes(x=year, y=count, 
                               group = parent_genres, color = parent_genres)) +
  geom_line()
```

There are some interesting patterns here! What do you see?

-   That green line is always on top

-   2020 crash in film numbers

-   2025 crash in film numbers

There is also a lot of information here, and it's a little hard to interpret information about the genres themselves with so many lines and the current color scheme.

## Changing the color palette

One visual tool is highlighting a single (or few) line(s) of interest. Say we were interested in trends in comedies. We could make all the other lines grey and leave comedies red to show them in context while highlighting the category of interest:

```{r}
ggplot(genre_summary_data, aes(x=year, y=count, group = parent_genres, color = parent_genres)) +
  geom_line(color = "grey", alpha = 0.5) + 
  geom_line(data = subset(genre_summary_data, parent_genres == "comedy"), color = "red")
```

What if you wanted to use a different color palette for your figure? Say cool tones:

```{r}
ggplot(genre_summary_data, aes(x=year, y=count, group = parent_genres, color = parent_genres)) +      
  geom_line() +       
  scale_color_hue(h = c(180, 300))
```

Or to explicitly control the color of each line? Maybe you've selected a color scheme to use for each throughout your paper, or there's a common convention in your field that you want to use.

```{r}
ggplot(genre_summary_data, aes(x=year, y=count, group = parent_genres, color = parent_genres)) +      
  geom_line() +       
  scale_color_manual(values=c("#69b3a2", "purple", "pink","red", "orange", "black","darkgrey", "lightblue", "yellow","violet", "lavender",  "white","darkgreen", "green","orchid","cyan" ))
```

There are good resources for your color adventures at <https://r-graph-gallery.com/ggplot2-color.html>, <https://ggplot2-book.org/scales-colour.html>, and <https://r-graph-gallery.com/package/paletteer.html>.

# (Stacked) barplots

Another way of looking at the genre-by-year data is called a stacked barplot. Instead of showing each genre as a line, this shows it as a part of a bar representing the year's total count.

First, let's do a simple barplot of the number of films\* per year. We need to calculate that yearly count (since we broke it down by parent_genres before):

*Note: This isn't actually the number of films - remember that each film can have multiple genres, which translates to multiple entries in this dataset.*

```{r}
year_summary <- movies |>   
  group_by(year) |>   
  summarise(     
    count = n()   
    )
```

Now let's graph that:

```{r}
ggplot(year_summary,        
       aes(x=year, y=count)) +   
  geom_bar(stat = "identity") 
```

Now let's split up those bars by the genre counts:

```{r}
stack <- ggplot(genre_summary_data, 
       aes(x=year, y=count, fill=parent_genres)) + # add fill option here
    geom_bar(position="stack", stat="identity")  # add stack option here

stack
```

We could use our manual color palette from above:

```{r}
stack + 
  scale_fill_manual(values=c("#69b3a2", "purple", "pink","red", "orange", "black","darkgrey", "lightblue", "yellow","violet", "lavender",  "white","darkgreen", "green","orchid","cyan" ))
```

## Themes

ggplot also has built-in themes that change the underlying appearance of figures. Let's look at some histograms of movie duration with various themes:

```{r}
ggplot(data = movies, mapping = aes(x=duration_minutes)) +      
  geom_histogram(binwidth=10) +      
  xlim(50,max(movies$duration_minutes))  
```

```{r}
ggplot(data = movies, mapping = aes(x=duration_minutes)) +      
  geom_histogram(binwidth=10) +      
  xlim(50,max(movies$duration_minutes)) +
  theme_classic()
```

```{r}
ggplot(data = movies, mapping = aes(x=duration_minutes)) +      
  geom_histogram(binwidth=10) +      
  xlim(50,max(movies$duration_minutes)) +
  theme_bw()
```

See built-in theme options here: <https://ggplot2.tidyverse.org/reference/ggtheme.html>, and a package offering extended options here: <https://r-graph-gallery.com/package/hrbrthemes.html>

# Ridgeline plots

Expanding on our duration histograms above and thinking back to genres - does length vary by genre? This is a lot of histograms to view, especially arranged like this:

```{r}
ggplot(data = movies_genre, 
       mapping = aes(x = duration_minutes)) +
  geom_histogram() +
  facet_wrap(~parent_genres)
```

An alternative plot type that allows us to stack many histograms is called a Ridgeline plot. To create these we need the `ggridges` package

```{r}
#install.packages("ggridges") 
library(ggridges)
```

```{r}
ggplot(movies_genre, 
       aes(x=duration_minutes, y= fct_rev(parent_genres), fill = parent_genres)) +  
  # fct_rev here reverses the factor labels so genres are arranged alphabetically going down the plot
  geom_density_ridges() +
  xlim(60,220) +
  theme(legend.position = "none") # remove the legend, which just repeats the Y axis
```

# Saving plots

`ggsave` is the function for saving ggplots. It defaults to saving the last plot displayed, at a default size and resolution depending on the filetype

```{r}
ggsave("../results/plot_default.png")
```

```{r}
ggsave("../results/plot_20x20cm.pdf", width = 20, height = 20, units = "cm")
```

```{r}
ggsave("../results/plot_highres.jpeg", dpi=600)
```

You can also assign `ggsave()` a specific plot to save:

```{r}
fig_one <- ggplot(data = movies, mapping = aes(x=duration_minutes)) +      
  geom_histogram(binwidth=10, fill ="red") +      
  xlim(50,max(movies$duration_minutes)) +
  theme_bw()


ggsave("../results/figure_one.png", fig_one)
```

# Reflection

-   What other questions do you have about this data that can be answered by a visual?

    -   What about movie duration vs rating?

-   How do you want to apply what you learned to your own data?
