---
title: "Other Visualizations"
author: "Katie Owers Bonner"
format: html
editor: visual
---

# Session 4: Additional visualizations

# ! material to add !

**Publishing Visualizations (30 mins?)**

1.  exporting visualizations, refining them for publication

    1.  save at higher resolution, in different formats

    2.  ggsave

2.  legends?

## Load data & packages if needed

```{r}
library(tidyverse)
library(here)
```

```{r}
df <- read_csv("../data/movie_data_dollars2.csv")
```

## The Question

Let's come back to Question A from yesterday: (**How) do critical and popular opinions about movies diverge?**

Our relevant variables are:

-   rating: IMDB's popular vote score (note the PG/R/etc. rating is called "mpa" in this dataset)

-   meta_score: Weighted average of professional critic scores

Reminder of what these variables look like from our exploratory analysis:

```{r}
summary(df[, c("rating", "meta_score")])
```

```{r}
ggplot(data =df, mapping = aes(x = meta_score, y=rating)) +
  geom_point() + 
  labs(
  x = "meta_score (critics)",
  y = "rating (the public)"
  )
```

## Does this association vary by MPA Rating?

Let's use an option called facet_wrap that splits up our plot by a factor in the data. Let's use the other kind of rating - MPA.

```{r}
ggplot(data =df, mapping = aes(x = meta_score, y=rating)) +
  geom_point()  + 
  labs(
  x = "meta_score (critics)",
  y = "rating (the public)"
  ) +
  facet_wrap(~mpa)
```

## What about splitting by genre?

Let's load in the file where each movie is split out by genre and they are categorized into parent genres:

```{r}
df_genre <- read_csv("../data/movie_split-by_genre.csv")

```

```{r}
df_genre |>
  summarise(unique_genres = n_distinct(genre),
            unique_parent_genres = n_distinct(parent_genre))

```

```{r}
table(df_genre$parent_genre)
```

And now do a figure where we facet by parent_genre:

```{r}
ggplot(data =df_genre, mapping = aes(x = meta_score, y=rating)) +
  geom_point() + 
  labs(
  x = "meta_score (critics)",
  y = "rating (the public)"
  )  +
  facet_wrap(~parent_genre)
```

# Boxplot/violin plot

What if we wanted to look instead at the values of rating (the public score) by parent_genre?

```{r}
ggplot(df_genre, mapping = aes(x = parent_genre, y = rating)) +
  geom_boxplot() + 
  theme(
  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )
```

What if we wanted to be able to see the point data as well as the summaries?

```{r}
ggplot(df_genre, mapping = aes(x = parent_genre, y = rating)) +
  geom_boxplot(fill = "darkgrey") +
  theme_minimal() +
  geom_jitter(color="blue", size=0.4, alpha=0.3) + theme(
  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
)
```

You can also sort the data so that the histograms are arranged according to their median value:

```{r}
ggplot(df_genre, mapping = aes(x = fct_reorder(parent_genre, -rating, .fun=median), y = rating)) +   
  geom_boxplot() +    
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)   )
```

What about another way of looking at this data? Violin plots show the shape of the data distribution, combining elements of histograms and the underlying points.

```{r}
ggplot(df_genre, mapping = aes(x = parent_genre, y = rating)) +
  geom_violin() + 
  theme(
  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )
```

# Line plots

What if you were curious about the number of movies in each parent genre by year?

First, let's calculate those yearly summaries we'll need:

```{r}
genre_summary_data <- df_genre |>
  group_by(year, parent_genre) |>
  summarise(
    count = n()
  )
```

```{r}
ggplot(genre_summary_data, aes(x=year, y=count, 
                               group = parent_genre, color = parent_genre)) +
  geom_line()
```

What if you wanted a different color palette?

```{r}
ggplot(summary_data, aes(x=year, y=count, group = parent_genre, color = parent_genre)) +
  geom_line() + 
  scale_color_hue(h = c(180, 300))
```

```{r}
ggplot(summary_data, aes(x=year, y=count, group = parent_genre, color = parent_genre)) +
  geom_line() + 
  scale_color_manual(values=c("#69b3a2", "purple", "pink","red", "orange", "black","darkgrey", "lightblue", "yellow","violet", "lavender",  "white","darkgreen", "green","orchid","cyan" ))
```

There are some good resources for your color adventures at <https://r-graph-gallery.com/ggplot2-color.html>!

Make this a ridgeline?by mean of the genre? % of total?

highlight a single line

Next steps

-   total number of films per year

-   total \# of genres per year

# Barplot (incl stacked)

Percent of these movie 4 genres over time (of the total)?

## Stacked barchart of genre by year

## Duration

```{r}
ggplot(data = df, mapping = aes(x=duration_minutes)) +   
  geom_histogram(binwidth=10) +   
  xlim(50,max(df$duration_minutes))   +   
  theme_classic()
```

```{r}
ggplot(data = df, mapping = aes(x=duration_minutes)) +   
  geom_density() +   
  xlim(50,max(df$duration_minutes))   +   
  theme_classic()
```

## 

# Aesthetics: Set ggplot Theme and Color Palette

# 

## Additional ideas to explore

-   What about movie duration vs rating?

# Reflection section

-   What other questions do you have about this data that can be answered by a visual?

-   Do you want to apply what you learned to your own data?
