---
title: "Clean Money Columns"
author: "Jeremy Mikecz"
format: html
editor: visual
---

# Clean Currency & Dates Columns

This notebook cleans and wrangles some troublesome columns into more useful formats:

-   currency columns (see below) are

```{r}
# install.packages("tidyverse")
library(tidyverse)
```

```{r}
df <- read_csv("../data/movie_data_2010-2025_clean_subset.csv")
```

```{r}
spec(df)
```

```{r}
df |>
  select(budget, gross_worldwide) |>
  slice_sample(n=30)
```

```{r}
colnames(df)
```

```{r}
tokens <- df |> 
  pull(budget) |>
  str_extract_all("[^0-9,. ]+") |>
  unlist() |>
  tibble(token = _) |>
  count(token, sort = TRUE)
```

```{r}
films_clean <- df |>
  mutate(
    budget_estimated = str_detect(budget, fixed("(estimated)")),
    budget_value = budget |>
      str_remove_all("[^0-9.]") |>
      as.numeric(),
    budget_currency = budget |>
      str_extract("[^0-9,. (estimated)]+") |>
      str_trim()
  )
```

Do the same thing but with the "gross_worldwide" column.

```{r}
films_clean <- films_clean |>
  mutate(
    grossww_value = gross_worldwide |>
      str_remove_all("[^0-9.]") |>
      as.numeric(),
    # grossw_estimate = str_detect(gross_worldwide, fixed("(estimated)")),
    grossww_currency = gross_worldwide |>
      str_extract("[^0-9,. (estimated)]+") |>
      str_trim()
  )
```

Double-check to ensure `NA`s were reads as NAs and not currencies.

```{r}
films_clean |>
  filter(is.na(budget_currency), !is.na(budget)) |>
  select(budget, budget_value, budget_currency) |>
  head(20)
```

`lubridate` is loaded automatically with `library(tidyverse)` and `mdy()` is tailor-made for the "Month Day, Year" format you have — it handles messy variations like missing leading zeros or abbreviated month names without complaint.

`yday()` handles leap years correctly by design — March 1st will be day 61 in a leap year and day 60 in a non-leap year, so February 29th naturally slots in as day 60 when it exists.

```{r}
films_clean <- films_clean |>
  mutate(
    # parse "June 18, 2010" → proper Date object
    date   = mdy(release_date) |> coalesce(ymd(str_c(release_date, "-01-01"))),
    #date = mdy(release_date) |> coalesce(ymd(release_date)),
    year   = year(date),
    month  = month(date),
    day    = day(date),
    daynum = yday(date),      # day of year, 1–366
    date_partial  = is.na(mdy(release_date)) & !is.na(date),  # TRUE = year-only entry
  )


```

```{r}
films_clean <- films_clean |>
  select(-any_of(c("budget", "gross_opening_wknd", "gross_worldwide", "gross_US_Canada")))
```

```{r}
write_csv(films_clean, "../data/movie_data_2010-2025_clean2.csv")
```

```{r}
films_clean_subset <- films_clean |>
  filter(budget_currency == "$" & grossww_currency == "$")

write_csv(films_clean_subset, "../data/movie_data_2010-2025_clean2_dollars.csv")
```

### Troubleshooting Dates

```{r}
films_clean |> 
  select(release_date, date, year, month, day, date_partial) |> 
  sample_n(20)
```

```{r}
films_clean |>
  filter(is.na(date), !is.na(release_date)) |>
  select(release_date) |>
  distinct()
```

```{r}
films_clean |>
  filter(is.na(date), !is.na(release_date)) |>
  select(release_date) |>
  distinct()
```
