---
title: "Intro to GGplot"
author: "Jeremy Mikecz"
format: html
editor: visual
---

## Preface: Choosing a plot type to answer a question

You may recall, one of our main questions is: what genre(s) of movies provide the best return on investment? In notebook 2, we ran some basic data analysis to arrive at an answer. However, now we turn to data visualization to see what else we can learn about the problem.

Before we start visualizing, we should consider what type of plot to create. If our question involves analyzing which genres of films produce the greatest ROI (profit over budget or expenditures), what plot types can best help us answer that question. Let's review [Data to Viz](https://www.data-to-viz.com) and [R Graph Gallery](https://r-graph-gallery.com/).

```         
```

But, what if you want to mix some elements of different plots? Create a plot type entirely? Data visualization experts and software developers recognized that to enhance customization, we should not confine ourselves to preset plot types. Rather, we should view data visualization as a languages with its own vocabulary and grammar.

Leland Wilkinson defined the "[Grammar of Graphics](https://link.springer.com/chapter/10.1007/0-387-28695-0_1) - in a book with that name - as a way to identify the smallest reproducible components of data visualizations. This **grammar of graphics** inspired the creators of R's most popular data visualization package **ggplot**, where the "gg" stands for "grammar of graphics".

\[more\]

## 0. Setup

First, let's import the packages we need:

```{r}
library(tidyverse)
library(here)
```

## 1. Import Data

Now let's import this dataset:

```{r}
movies <- read_csv(here("data/movie_data_financial.csv"))
```

```{r}
colnames(movies)
```

```{r}

```

## 2. First Plot: two variables

### 2a. Basic plot

Let's first create a scatterplot of this dataset for one year (1982) and two variables (`lifeExp` = life expectancy and `gdpPercap` = gdpPercap).

```{r}
```

```{r}
# let's store the year we want to examine in variable 
# so we can easily change it later
ggplot(movies, aes(x=budget_mm, y=grossww_mm)) +
  geom_point()
```

### 2b. Adjust color & transparency (as constants)

[ggplot colors](https://sape.inf.usi.ch/quick-reference/ggplot2/colour)

```{r}
ggplot(movies, aes(x=budget_mm, y=grossww_mm)) +
  geom_point(color="steelblue4", alpha = 0.3, size=3)
```

### 2c. Adjust color as a variable

```{r}
ggplot(movies, aes(x=budget_mm, y=grossww_mm, color=mpa)) +
  geom_point(alpha = 0.3, size=3)
```

### 2c. Add a smoothed line

```{r}
p <- ggplot(movies, aes(x=budget_mm, y=grossww_mm)) +
  geom_point(color="steelblue4", alpha = 0.3, size=3) +
  geom_smooth(color="purple")                      # , method='lm', se=False

p
```

### 2 - Extra: examine distributions more closely

```{r}
#install.packages("ggExtra")
library(ggExtra)

ggMarginal(p, type='histogram')

```

## Dealing with Small Budget Movies: Rescale Data

```{r}
library(scales)
log_p <- ggplot(movies, aes(x=budget_mm, y=grossww_mm)) +
  geom_point(color="steelblue4", alpha = 0.3, size=3) +
  geom_smooth(color="purple") +
  scale_x_log10(labels = label_dollar(suffix = "M")) +
  scale_y_log10(labels = label_dollar(suffix = "M")) +
  annotation_logticks()
  
log_p
```

## 3. Working with Genres

```{r}
genre_df <- read_csv(here("data/movies_split-by_parent-genre.csv"))
```

```{r}
head(genre_df)
```

```{r}
genre2020_df <- genre_df |>
  filter(year >= 2020)
```

```{r}
ggplot(genre2020_df, aes(x=budget_mm, y=grossww_mm, color=parent_genres)) +
  geom_point(alpha = 0.3, size=3) +
  geom_smooth(color="purple") +
  scale_x_log10(labels = label_dollar(suffix = "M")) +
  scale_y_log10(labels = label_dollar(suffix = "M")) +
  annotation_logticks() 
  
```

```{r}
pgenres <- c("comedy",
             "drama", 
             "romance")
for (pgenre in pgenres) {
  print(pgenre)
  print(
    ggplot(genre2020_df, aes(x=budget_mm, y=grossww_mm)) +
      geom_point(size=1, color="gray", alpha=0.4) +
      geom_point(data=genre2020_df |> filter(parent_genres==pgenre),
                 color="blue", size=2, alpha = 0.6) +
      geom_smooth(color="purple") +
      scale_x_log10(labels = label_dollar(suffix = "M")) +
      scale_y_log10(labels = label_dollar(suffix = "M")) +
      annotation_logticks() 
  )
}
```

```{r}
ggplot(genre2020_df, aes(x = budget_mm, y = grossww_mm)) +
  geom_point(alpha = 0.3, size = 3) +
  geom_smooth(color="purple") +
  scale_x_log10(labels = label_dollar(suffix = "M")) +
  scale_y_log10(labels = label_dollar(suffix = "M")) +
  annotation_logticks() +
  facet_wrap(~ parent_genres, ncol=2)
```

```{r}
colnames(genre2020_df)
```

```{r}
ggplot(genre2020_df, aes(x = budget_mm, y = grossww_mm)) +
  geom_point(alpha = 0.3, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  facet_wrap(~ parent_genres, ncol=3) +
  labs(
    title    = "Budget vs. Worldwide Gross by Genre",
    subtitle = "Dashed line = break-even",
    x        = "Budget ($ millions)",
    y        = "Worldwide Gross ($ millions)"
  )
```

## ROI Distribution by Genre (Box Plot)

A bar chart of medians hides the spread. Use a box plot to see the **full distribution** of ROI across genres, including variability and outliers.

```{r}
ggplot(genre2020_df, aes(x = reorder(parent_genres, profit_pct, median), y = profit_pct)) +
  geom_boxplot(outlier.alpha = 0.2, outlier.shape = NA) +
  geom_jitter(size=0.4, width=0.3, alpha=0.3, aes(color=parent_genres)) +
  coord_flip() +
  labs(
    title = "Distribution of ROI by Genre",
    x     = "Genre",
    y     = "ROI (%)"
  )

ggsave(here("results/movie-genres_boxplot-roi.png"))
```

## High-Return Low-Budget Films (Scatter Plot, Interactive)

To directly highlight films that made a lot relative to a small budget, we can size or color points by `profit_pct`, and filter to a budget range of interest. Using **plotly** makes it easy to hover and identify specific films.

```{r}
# install.packages("plotly")
library(plotly)

low_budget <- genre2020_df |>
  filter(budget_mm <= 50, profit_pct > 0)

p <- ggplot(low_budget, aes(
    x      = budget_mm,
    y      = grossww_mm,
    color  = parent_genres,
    size   = profit_pct,
    text   = paste0(title, "\nROI: ", round(profit_pct), "%")
  )) +
  geom_point(alpha = 0.5) +
  labs(
    title = "Low-Budget Films (≤ $50M) by Worldwide Gross",
    x     = "Budget ($ millions)",
    y     = "Worldwide Gross ($ millions)"
  )

ggplotly(p, tooltip = "text")
```

## Risk-Return Scatterplot

```{r}

median_ROI_overall <- median(genre2020_df$profit_pct)
print(median_ROI_overall)
mean(genre2020_df$grossww_mm > 5 * genre2020_df$budget_mm)

```

```{r}

```

```{r}
genre_summary <- genre2020_df |>
  group_by(parent_genres) |>
  summarize(
    median_roi = median(profit_pct, na.rm = TRUE),
    hit_rate   = mean(grossww_value > 5 * budget_value, na.rm = TRUE),
    n          = n()
  ) |>
  filter(n >= 50)
```

```{r}
# install.packages("ggrepel")
library(ggrepel)

ggplot(genre_summary, aes(x = median_roi, y = hit_rate, size = n, label = parent_genres)) +
  geom_point(alpha = 0.7, color = "steelblue") +
  geom_text_repel(size = 3) +          # from ggrepel package
  scale_y_continuous(labels = scales::percent) +
  # scale_x_continuous(labels = scales::percent) +
  geom_hline(yintercept=.118, linetype="dashed") +
  geom_vline(xintercept=13, linetype="dashed") +
  labs(
    title = "Which genres are worth the bet?",
    x = "Median ROI", y = "Hit Rate (revenue > 5× budget)"
  ) +
  theme_minimal()

```

## Animated Scatter

```{r}
library(plotly)

p <- genre2020_df |>
  filter(!is.na(budget_value), !is.na(grossww_value), !is.na(year)) |>
  ggplot(aes(x = budget_value, y = grossww_value, 
             color = parent_genres, frame = year,
             text = paste("Title:", title, "<br>Genre:", parent_genres))) +
  geom_point(alpha = 0.6, size = 2) +
  geom_abline(slope = 5, intercept = 0,   # the "hit" threshold line
              linetype = "dashed", color = "black") +
  scale_x_log10(labels = scales::dollar_format()) +
  scale_y_log10(labels = scales::dollar_format()) +
  labs(title = "Budget vs. Worldwide Gross by Genre ({frame_time})",
       x = "Production Budget", y = "Worldwide Gross") +
  theme_minimal()

ggplotly(p, tooltip = "text") |>
  animation_opts(frame = 1000, redraw = FALSE)
```

## Diverging Bar Chart

```{r}

films_filtered <- genre2020_df |>
  filter(!is.na(profit_pct), !is.na(parent_genres), parent_genres != "?",
         !is.na(grossww_value), !is.na(budget_value)) |>
  group_by(parent_genres) |>
  filter(n() >= 50) |>
  ungroup()

genre_order <- films_filtered |>
  group_by(parent_genres) |>
  summarize(hit_rate = mean(grossww_value > 5 * budget_value, na.rm = TRUE)) |>
  arrange(hit_rate) |>
  pull(parent_genres)  

films_filtered|>
  mutate(outcome = case_when(
    grossww_value > 5 * budget_value  ~ "Hit (>5× budget)",
    grossww_value < budget_value      ~ "Flop (<1× budget)",
    TRUE                              ~ "Middle"
  )) |>
  group_by(parent_genres, outcome) |>
  summarize(n = n()) |>
  mutate(pct = n / sum(n),
         parent_genres = factor(parent_genres, levels=genre_order)) |>
  ggplot(aes(x = parent_genres, y = pct, fill = outcome)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("Hit (>5× budget)" = "#2ecc71", 
                                "Middle"           = "#bdc3c7",
                                "Flop (<1× budget)"= "#e74c3c")) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Outcomes by genre", x = NULL, y = "Share of films") +
  theme_minimal()
```

-   

-   theme_minimal? basic preset themes

-   clusters?

-   aggreg by genre:

    -   animate by year

    -   hover

    -   connected scatter plot
