---
title: "Wrangle and Analyze Film Data"
author: "Jeremy Mikecz"
format: html
editor: visual
---

```{r}
library(tidyverse)
```

```{r}
df_temp <- read_csv("../../data/movie_data_2010-2025_clean_dollars.csv")
```

```{r}
library(here)
df <- read_csv(here("data/movie_data_2010-2025_clean_dollars.csv"))
```

```{r}
head(df)
```

Opener and Closer - referring back to big research questions

## Standardize Variable Names

## Clean / Wrangle Scores

1.  01 - look at list of cleaning steps as is....

## remove extreme or suspicious values

## interactive portions

# Calculate Profit

## Make Numbers more manageable

```{r}
summary(df$grossww_value)
```

## Visualize Distributions

Convert to millions:

```{r}
df <- df |> 
  mutate(budget_mm = budget_value / 1000000,
         grossww_mm = grossww_value / 1000000)
```

Calculate Profit

```{r}
df = df |>
  mutate(net_profit_mm = grossww_mm - budget_mm)
```

Calculate Profit as Percent of Budget

```{r}
df = df |>
  mutate(profit_pct = net_profit_mm /  budget_mm * 100)
```

## 

# Review Genres

Let's take a look at the genre column:

```{r}
df$genres[0:10]
```

Each film appears once per genre, so multi-genre films are counted multiple times. This means genres that frequently co-occur may have correlated estimates. It also means a film with two assigned genres is counted just twice while one with five is counted five times.

Some possible solutions:

-   **one row per film/genre (EASIEST)**: split genres into different rows, so that a five-genre film is split out into five rows

-   **one row per film genre (weighted)**: split genres into different rows but weight each genre by its proportion. So a "horror-comedy" means horror counts for 0.5 genres and comedy also for 0.5.

-   **one row per film, one column per genre**\*\* (wide/dummy format): create a separate binary column (TRUE / FALSE) for each genre; best for statistical modeling and controlling (i.e. "controlling for other genres, what is horror's effect?")

Let's do the simplest approach: one row per film / genre (no weighting).

*whiteboard*

*demo string functions a bit*

```{r}
genre_df <- df |>
  mutate(
    genre = genres |>
      str_remove_all("\\[|\\]|'") |>   # remove brackets and single quotes
      str_trim()                         # remove any leading/trailing whitespace
  ) |>
  separate_rows(genre, sep = ",") |>   # split on comma
  mutate(genre = str_trim(genre))     # clean up whitespace around each genre name
```

### Verify

```{r}
genre_df |> 
  filter(str_detect(genre, '"')) |> 
  select(genre) |> 
  head()
```

## Examine Distributions of Genres

```{r}

```

## Median ROI by genre

```{r}
genre_aggreg <- genre_df |>
  group_by(genre) |>
  summarize(
    median_roi = median(profit_pct, na.rm = TRUE),
    n = n()
  ) |>
  arrange(desc(median_roi))
  
```

```{r}
genre_aggreg |>
  filter (n>=20)
```

**Add spread to the summary** Median alone doesn't capture risk. A genre with high median ROI but huge variance is a very different bet than one with moderate but consistent returns:

```{r}
genre_df |>
  group_by(genre) |>
  summarize(
    median_roi = median(profit_pct, na.rm = TRUE),
    iqr_roi    = IQR(profit_pct, na.rm = TRUE),   # interquartile range = spread
    n          = n()
  ) |>
  filter(n >= 50) |>
  arrange(desc(median_roi))
```

The IQR is a natural companion to the median and tells you about consistency — which genres are reliable bets vs. high-risk/high-reward.

```{r}

```

## 

```{r}

```

```{r}

```

```{r}

```
