---
title: "Setup"
author: "Jeremy Mikecz"
date: "2026-01-07"
format: html
editor: visual
---

------------------------------------------------------------------------

# Setup Instructions

This document provides detailed instructions for setting up your R environment for this course/workshop.

## Table of Contents

1.  [Installing R and RStudio](#installing-r-and-rstudio)
2.  [Downloading Course Materials](#downloading-course-materials)
3.  [Setting Up the R Environment](#setting-up-the-r-environment)
4.  [Verifying Your Setup](#verifying-your-setup)
5.  [Troubleshooting](#troubleshooting)

## Installing R and RStudio {#installing-r-and-rstudio}

### Install R

1.  Go to <https://posit.co/download/rstudio-desktop/>
2.  Select `Download and Install R`
3.  Click on the link for your operating system (Windows, Mac, or Linux)
4.  Download and install the latest version (4.0 or higher)

**Windows users:** Download "base" R\
**Mac users:** Choose the appropriate .pkg file for your chip (Intel or Apple Silicon)\
**Linux users:** Follow distribution-specific instructions

### Install RStudio (Recommended)

1.  Go to <https://posit.co/download/rstudio-desktop/>
2.  Download RStudio Desktop (free version)
3.  Install using the downloaded installer

**Alternative:** You can use other R interfaces, but course instructions assume RStudio.

## Downloading Course Materials {#downloading-course-materials}

### Option 1: Download as ZIP

*This method is recommended for newer programmers not yet familiar with git and Github and for any users who only needs a static copy of the repository.*

1.  open the Github course repository.
2.  Click the green "Code" button on the repository page
3.  Select "Download ZIP"
4.  Extract the ZIP file to a memorable location (e.g., \~/Documents/R-Courses/)
5.  Important: To get updates, you'll need to re-download

## Setting Up the R Environment (optional) {#setting-up-the-r-environment}

This repository uses renv for package management, ensuring everyone has the same package versions.

### 1. Open the Project

Double-click the \[project-name\].Rproj file to open the project in RStudio.

### 2. Setup Project Folder

### 3. Install renv

If prompted to install renv, click "Yes" or run:

```{r}
install.packages("renv")
```

### 4. Initiate `renv` and load packages

To initiate a virtual environment for your project, you have two options, depending if a) you are creating a new project, or b) importing an existing project.

**The instructions below demonstrate how to create a new project (option 1 above).** For instructions for option 2, see **NOTEBOOK ??**.

#### 4A. Initiate a new `renv` virtual environment for your project.

First, you will need to initiate a new virtual environment for this project, by running the following in the R console:

```         
renv::init()
```

### 5. Write first .R script

### 6. Write first Quarto Document (notebook)

### Step 4: Install a new R package to your virtual environment

For `renv`, you can still install packages the traditional way, using `install.packages("package-name")`. However, it is recommended that you use `renv::install()` to install packages. You can run the code cell below or run it in the R console (note: when installing packages, frequently the console will prompt you to confirm if you want to proceed, so even if you install from a notebook or script you may want to check to console to see if an unanswered prompt is holding your code up).

```{r}
renv::install("here")
```

### Step 5: Review a list of installed packages

To output a list of installed packages, uncomment out and run the code below:

```{r}
# installed.packages()
```

### Step 6: Updating your `renv.lock` file

As you add new packages to your local virtual environment, periodically you'll want to update the `renv.lock` file so that includes the latest and most complete list of all installed packages and their versions. You can check to see if the **renv library** (where packages are installed, found in `renv/library` ) and the **renv lockfile** (`renv.lock`) are in sync.

```{r}
renv::status()
```

If they are not, you can update the lockfile with:

```{r}
renv::snapshot()
```

On the other hand, if you are importing a project from someone or somewhere else, you can install all dependencies listed in the lockfile to the library using `renv::restore()`. Review the [renv documentation](https://rstudio.github.io/renv/articles/renv.html) for more information.

## 
