---
title: "Setup"
author: "Jeremy Mikecz"
date: "2026-01-07"
format: html
editor: visual
---

------------------------------------------------------------------------

# Setup Instructions

This document provides detailed instructions for setting up your R environment for this course/workshop.

## Table of Contents

1.  [Installing R and RStudio](#installing-r-and-rstudio)
2.  [Downloading Course Materials](#downloading-course-materials)
3.  [Setting Up the R Environment](#setting-up-the-r-environment)
4.  [Verifying Your Setup](#verifying-your-setup)
5.  [Troubleshooting](#troubleshooting)

## Installing R and RStudio {#installing-r-and-rstudio}

### Install R

1.  Go to <https://posit.co/download/rstudio-desktop/>
2.  Select `Download and Install R`
3.  Click on the link for your operating system (Windows, Mac, or Linux)
4.  Download and install the latest version (4.0 or higher)

**Windows users:** Download "base" R\
**Mac users:** Choose the appropriate .pkg file for your chip (Intel or Apple Silicon)\
**Linux users:** Follow distribution-specific instructions

### Install RStudio (Recommended)

1.  Go to <https://posit.co/download/rstudio-desktop/>
2.  Download RStudio Desktop (free version)
3.  Install using the downloaded installer

**Alternative:** You can use other R interfaces, but course instructions assume RStudio.

## Downloading Course Materials {#downloading-course-materials}

### Option 1: Download as ZIP

*This method is recommended for newer programmers not yet familiar with git and Github and for any users who only needs a static copy of the repository.*

1.  open the Github course repository.
2.  Click the green "Code" button on the repository page
3.  Select "Download ZIP"
4.  Extract the ZIP file to a memorable location (e.g., \~/Documents/R-Courses/)
5.  Important: To get updates, you'll need to re-download

### Option 2: Using Git (Advanced)

\*This method is recommended for users familiar with git and Github, have git installed, have a Github account, and who wish to create a dynamic copy ("clone") of the repository. A clone allows you to update your local repo when its online counterpart as been updated (using the `pull` command in git). It also allows you to push any changes you made locally back up to the remote repository, but only if the repository administrator has given you these privileges.\*

If you have Git installed:

1.  open the Github course repository.
2.  If you have an SSH key set up for your Github account: Select the Green `Code` Button –\> SSH –\> copy the SSH URL
3.  If you do not have an SSH key set up for your Github account: Select the Green `Code` Button –\> HTTPS –\> copy the HTTPS URL
4.  In Git Bash, navigate to a memorable folder where you want to clone the course repository (i.e. \~/Documents/R-Courses/). Then type in the following (using the REPOSITORY SSH or HTTPS URL you copied in the previous steps.

In a terminal:

```         
git clone [REPOSITORY-URL]
cd [repository-name]
```

To update materials throughout the course:

```{bash}
git pull origin main

```

## Setting Up the R Environment {#setting-up-the-r-environment}

This repository uses renv for package management, ensuring everyone has the same package versions.

### Step 1: Open the Project

Double-click the \[project-name\].Rproj file to open the project in RStudio.

### Step 2: Install renv

If prompted to install renv, click "Yes" or run:

```{r}
install.packages("renv")
```

### Step 3: Initiate `renv` and load packages

To initiate a virtual environment for your project, you have two options, depending if:

1.  you are working with a new project OR you are working with an existing project with no `renv.lock`. Follow Step 3A below to initiate a new renv virtual environment.
2.  You are working with an existing project containing a list of dependencies for the project in a `renv.lock` file found in the project's root directory. Follow Step 3B below.

There is a third possibility. You may find yourself working with an existing project that does have a list of dependencies, but store in an alternate format (i.e. using the now deprecated R package `packrat`). If that's the case you will have to search online for a solution. If you are working with a packrat project, for example, the `renv` webpage contains instructions for converting a packrat project to a renv project.

#### Step 3A: Initiate a new `renv` virtual environment for your project.

If you are creating a new project or working with an existing project with no `renv.lock` file found in the project root directory, you will need to initiate a new virtual environment for this project, by running the following in the R console:

`{renv::init()}`

#### 3B: Load packages saved in existing `renv.lock` file

If you are working with an existing project containing a list of project dependencies in a `renv.lock` file (located in the project's root directory). In the R console, run:

`{renv::restore()}`

This will install all required packages. This may take 10-30 minutes depending on your system and internet connection.

Type `y` when prompted to proceed.

### Step 4: Install a new R package to your virtual environment

For `renv`, you can still install packages the traditional way, using `install.packages("package-name")`. However, it is recommended that you use `renv::install()` to install packages. You can run the code cell below or run it in the R console (note: when installing packages, frequently the console will prompt you to confirm if you want to proceed, so even if you install from a notebook or script you may want to check to console to see if an unanswered prompt is holding your code up).

```{r}
renv::install("here")
```

### Step 5: Review a list of installed packages

To output a list of installed packages, uncomment out and run the code below:

```{r}
# installed.packages()
```

### Step 6: Updating your `renv.lock` file

As you add new packages to your local virtual environment, periodically you'll want to update the `renv.lock` file so that includes the latest and most complete list of all installed packages and their versions. You can check to see if the **renv library** (where packages are installed, found in `renv/library` ) and the **renv lockfile** (`renv.lock`) are in sync.

```{r}
renv::status()
```

If they are not, you can update the lockfile with:

```{r}
renv::snapshot()
```

On the other hand, if you are importing a project from someone or somewhere else, you can install all dependencies listed in the lockfile to the library using `renv::restore()`. Review the [renv documentation](https://rstudio.github.io/renv/articles/renv.html) for more information.

## Test the Setup

Run the setup verification script:

```{r}
source("scripts/00_setup.R")
```

If you see "✓ Setup complete!" you're ready to go!

## Verifying Your Setup {#verifying-your-setup}

Run this code in your R console to verify everything is working:

```{r}
# Check R version
R.version.string

# Check if key packages load
required_packages <- c("tidyverse", "rmarkdown", "here")
sapply(required_packages, require, character.only = TRUE)

# Test plot generation
library(ggplot2)
ggplot(mtcars, aes(x = mpg, y = wt)) +
  geom_point() +
  labs(title = "Test Plot - If you see this, ggplot2 works!")
```

## Troubleshooting {#troubleshooting}

### Issue: "renv::restore() fails with compilation errors"

#### Solution 1: Install binary packages instead of compiling from source

```{r}
# On Mac/Linux, when prompted about installing from source, choose "No"
# Or set this option:
options(install.packages.check.source = "no")
renv::restore()
```

#### Solution 2: Install system dependencies (Linux users)

On Ubuntu/Debian:

```{bash}
sudo apt-get install libxml2-dev libcurl4-openssl-dev libssl-dev
```

### Issue: "Cannot open file 'data/raw/myfile.csv'"

#### Solution: Always use the here package for file paths

```{r}
library(here)
data <- read.csv(here("data", "raw", "myfile.csv"))
```

Using **setwd()** is not recommended \[EXPLAIN WHY!\]

### Issue: "Package 'X' is not available for R version Y"

#### Solution: Update R to the latest version, or use an older package version

```{r}
# Install specific older version
renv::install("package@version")
```

### Issue: RStudio doesn't recognize the project

#### Solution: Ensure you opened the `.Rproj` file, not just a script

-   Close RStudio

-   Navigate to the repository folder

-   Double-click the .Rproj file

### Issue: Git clone fails

#### Solution: Check your network connection, or try HTTPS instead of SSH

```{bash}
# If SSH fails, try HTTPS
git clone https://github.com/username/repository.git
```

### Still Having Issues?

\[insert possible steps, i.e. contact instructor, review particular online materials, etc.\]
